---
title: "Bayesian Surprise with Canadian Census Data (cancensus)"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Bayesian Surprise with Canadian Census Data (cancensus)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE,
  fig.alt = "Canadian Census example visualization of Bayesian surprise or funnel plot values."
)

is_pkgdown <- identical(Sys.getenv("IN_PKGDOWN"), "true")
has_cancensus_key <- nzchar(Sys.getenv("CM_API_KEY")) ||
  nzchar(Sys.getenv("CANCENSUS_API_KEY"))
run_cancensus_examples <- is_pkgdown && has_cancensus_key
```

# Introduction

This vignette demonstrates how to use the `bayesiansurprise` package with
Canadian Census data accessed through the `cancensus` package. The cancensus
package provides access to Canadian Census data from 1996 through 2021 via
the CensusMapper API.

Canadian Census data is a useful setting for Bayesian Surprise analysis because
it provides:

- Comprehensive population data at multiple geographic levels
- Rich demographic and socioeconomic variables
- Consistent geography across Census years
- Both counts and rates for many variables

## Prerequisites

You'll need a CensusMapper API key from https://censusmapper.ca/users/sign_up

```{r load-packages}
library(bayesiansurprise)
library(cancensus)
library(sf)
library(ggplot2)
library(dplyr)

# Set your CensusMapper API key (do this once)
# set_cancensus_api_key("YOUR_KEY_HERE", install = TRUE)

# Optionally set a cache path for faster subsequent queries
# set_cancensus_cache_path("path/to/cache", install = TRUE)
```

***

# Understanding cancensus Data Structure

Before diving into analysis, let's understand the available datasets and
variables.

```{r list-datasets, eval=FALSE}
# List available Census datasets
datasets <- list_census_datasets()
print(datasets %>% select(dataset, description))
```

```{r list-regions, eval=FALSE}
# List regions for the 2021 Census
regions_2021 <- list_census_regions("CA21")
head(regions_2021 %>% filter(level == "CMA"))
```

***

# Example 1: Low Income Rates by Census Division

Let's examine low income rates across Canada's Census Divisions and identify
which regions have high or low surprise under an explicit population and
sampling-variation model space.

## Fetch Census Data

```{r fetch-income-data, eval=run_cancensus_examples}
# Find relevant vectors for low income
# v_CA21_1085: Total - Low-income status in 2020
# v_CA21_1086: 0 to 17 years in low income
# v_CA21_1090: In low income

# Get low income data at the Census Division level for all of Canada
# Using labels = "short" for cleaner column names
income_data <- get_census(
  dataset = "CA21",
  regions = list(C = "01"),  # All of Canada
  vectors = c(
    total_pop = "v_CA21_1085",  # Total population for low income status
    low_income = "v_CA21_1090"   # Population in low income
  ),
  level = "CD",
  geo_format = "sf",
  labels = "short",
  quiet = TRUE
)

# Clean and filter
income_data <- income_data %>%
  filter(!is.na(total_pop) & !is.na(low_income)) %>%
  filter(total_pop > 0) %>%
  mutate(
    low_income_rate = low_income / total_pop
  )

head(income_data %>% st_drop_geometry() %>% select(GeoUID, `Region Name`, total_pop, low_income, low_income_rate))
```

## Compute Bayesian Surprise

```{r compute-income-surprise, eval=run_cancensus_examples}
# Compute surprise
income_surprise <- surprise(
  income_data,
  observed = low_income,
  expected = total_pop,
  models = c("uniform", "baserate", "funnel")
)

print(income_surprise)
summary(income_surprise)
```

## Visualize Surprising Regions

```{r plot-income-surprise, fig.width=10, fig.height=8, eval=run_cancensus_examples}
# Add surprise values to data
income_data$surprise <- get_surprise(income_surprise, "surprise")
income_data$signed_surprise <- get_surprise(income_surprise, "signed")

# Transform to Statistics Canada Lambert projection for better Canada visualization
income_data_lambert <- st_transform(income_data, "EPSG:3347")

# Plot Canada-wide
ggplot(income_data_lambert) +
  geom_sf(aes(fill = signed_surprise), color = "white", linewidth = 0.1) +
  scale_fill_surprise_diverging(limits = c(-1.5, 1.5), oob = scales::squish) +
  labs(
    title = "Surprising Low Income Rates by Census Division",
    subtitle = "Red = Higher than expected | Blue = Lower than expected",
    caption = "Source: Statistics Canada, 2021 Census"
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

## Identify Most Surprising Regions

```{r top-surprise-regions, eval=run_cancensus_examples}
# Top regions with positive signed surprise
high_income_surprise <- income_data %>%
  st_drop_geometry() %>%
  filter(signed_surprise > 0) %>%
  arrange(desc(surprise)) %>%
  select(`Region Name`, low_income_rate, total_pop, surprise) %>%
  head(10)

cat("Census Divisions with positive signed surprise:\n")
print(high_income_surprise)

# Top regions with negative signed surprise
low_income_surprise <- income_data %>%
  st_drop_geometry() %>%
  filter(signed_surprise < 0) %>%
  arrange(desc(surprise)) %>%
  select(`Region Name`, low_income_rate, total_pop, surprise) %>%
  head(10)

cat("\nCensus Divisions with negative signed surprise:\n")
print(low_income_surprise)
```

***

# Example 2: Housing Analysis in Metro Vancouver

Let's analyze owner-occupied housing rates at the Census Tract level within
the Vancouver CMA.

```{r fetch-housing-data, eval=run_cancensus_examples}
# Get housing tenure data for Vancouver CMA
# v_CA21_4237: Total private households by tenure
# v_CA21_4238: Owner households

housing_data <- get_census(
  dataset = "CA21",
  regions = list(CMA = "59933"),  # Vancouver CMA
  vectors = c(
    total_hh = "v_CA21_4237",  # Total households
    owner_hh = "v_CA21_4238"   # Owner households
  ),
  level = "CT",
  geo_format = "sf",
  labels = "short",
  quiet = TRUE
)

housing_data <- housing_data %>%
  filter(!is.na(total_hh) & !is.na(owner_hh)) %>%
  filter(total_hh > 50) %>%  # Filter small tracts
  mutate(
    owner_rate = owner_hh / total_hh
  )

cat("Number of Census Tracts:", nrow(housing_data), "\n")
```

```{r housing-surprise, fig.width=9, fig.height=7, eval=run_cancensus_examples}
# Compute surprise - use uniform and baserate only for local CMA analysis
# (funnel model absorbs too much variation at this scale)
housing_surprise <- surprise(
  housing_data,
  observed = owner_hh,
  expected = total_hh,
  models = c("uniform", "baserate")
)

housing_data$signed_surprise <- get_surprise(housing_surprise, "signed")

# Plot Vancouver
ggplot(housing_data) +
  geom_sf(aes(fill = signed_surprise), color = "white", linewidth = 0.02) +
  scale_fill_surprise_diverging(limits = c(-1.5, 1.5), oob = scales::squish) +
  labs(
    title = "Surprising Home Ownership Rates in Metro Vancouver",
    subtitle = "Census Tract Level | Red = More owners | Blue = Fewer owners than expected",
    caption = "Source: Statistics Canada, 2021 Census"
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 9)
  )
```

***

# Example 3: Language at Home Analysis

Identify areas with high model-conditional surprise for non-official language
speakers at home.

```{r fetch-language-data, eval=run_cancensus_examples}
# Get language data for Toronto CMA
# v_CA21_1144: Total - Knowledge of official languages
# v_CA21_1153: Neither English nor French

language_data <- get_census(
  dataset = "CA21",
  regions = list(CMA = "35535"),  # Toronto CMA
  vectors = c(
    total_pop = "v_CA21_1144",  # Total population
    no_official = "v_CA21_1153"   # Neither official language
  ),
  level = "CT",
  geo_format = "sf",
  labels = "short",
  quiet = TRUE
)

language_data <- language_data %>%
  filter(!is.na(total_pop) & !is.na(no_official)) %>%
  filter(total_pop > 100)
```

```{r language-surprise, fig.width=9, fig.height=7, eval=run_cancensus_examples}
# Compute surprise - use uniform and baserate only for local CMA analysis
lang_surprise <- surprise(
  language_data,
  observed = no_official,
  expected = total_pop,
  models = c("uniform", "baserate")
)

language_data$signed_surprise <- get_surprise(lang_surprise, "signed")

# Plot Toronto
ggplot(language_data) +
  geom_sf(aes(fill = signed_surprise), color = "white", linewidth = 0.02) +
  scale_fill_surprise_diverging(limits = c(-1.5, 1.5), oob = scales::squish) +
  labs(
    title = "Surprising Non-Official Language Concentrations in Toronto",
    subtitle = "Census Tract Level | Red = Higher | Blue = Lower concentration than expected",
    caption = "Source: Statistics Canada, 2021 Census"
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 9)
  )
```

***

# Example 4: Provincial Comparison with Funnel Plot

Compare provinces using a funnel plot to account for population size variation.

```{r provincial-funnel, eval=run_cancensus_examples}
# Get provincial data
provincial_data <- get_census(
  dataset = "CA21",
  regions = list(C = "01"),
  vectors = c(
    total_pop = "v_CA21_1085",  # Total for low income
    low_income = "v_CA21_1090"   # In low income
  ),
  level = "PR",
  geo_format = NA,
  labels = "short",
  quiet = TRUE
)

provincial_data <- provincial_data %>%
  filter(!is.na(total_pop) & total_pop > 0)

# Compute funnel data for points
funnel_df <- compute_funnel_data(
  observed = provincial_data$low_income,
  sample_size = provincial_data$total_pop,
  type = "count",
  limits = c(2, 3)
)
funnel_df$province <- provincial_data$`Region Name`
funnel_df$rate <- funnel_df$observed / funnel_df$sample_size

# National rate
national_rate <- sum(provincial_data$low_income) / sum(provincial_data$total_pop)

# Create continuous funnel bands
n_range <- range(provincial_data$total_pop)
n_seq <- exp(seq(log(n_range[1] * 0.8), log(n_range[2] * 1.2), length.out = 200))
se_seq <- sqrt(national_rate * (1 - national_rate) / n_seq)

funnel_bands <- data.frame(
  n = n_seq,
  rate = national_rate,
  lower_2sd = national_rate - 2 * se_seq,
  upper_2sd = national_rate + 2 * se_seq,
  lower_3sd = national_rate - 3 * se_seq,
  upper_3sd = national_rate + 3 * se_seq
)

# Create funnel plot
ggplot() +
  # 3 SD band (outer)
  geom_ribbon(
    data = funnel_bands,
    aes(x = n, ymin = lower_3sd, ymax = upper_3sd),
    fill = "#9ecae1", alpha = 0.5
  ) +
  # 2 SD band (inner)
  geom_ribbon(
    data = funnel_bands,
    aes(x = n, ymin = lower_2sd, ymax = upper_2sd),
    fill = "#3182bd", alpha = 0.5
  ) +
  # National average line
  geom_hline(
    yintercept = national_rate,
    linetype = "dashed", color = "#636363", linewidth = 0.8
  ) +
  # Points
  geom_point(
    data = funnel_df,
    aes(x = sample_size, y = rate, color = abs(z_score) > 2),
    size = 4
  ) +
  # Labels
  ggrepel::geom_text_repel(
    data = funnel_df,
    aes(x = sample_size, y = rate, label = province),
    size = 3, fontface = "bold",
    max.overlaps = 15,
    segment.color = "gray60"
  ) +
  scale_color_manual(
    values = c("FALSE" = "#636363", "TRUE" = "#e6550d"),
    guide = "none"
  ) +
  scale_x_log10(
    labels = scales::comma,
    breaks = c(1e4, 1e5, 1e6, 1e7)
  ) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Funnel Plot: Provincial Low Income Rates",
    subtitle = "Provinces outside bands have surprising rates given population size",
    x = "Population (log scale)",
    y = "Low Income Rate",
    caption = "Source: Statistics Canada, 2021 Census"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

***

# Example 5: Comparing Census Years

Compare surprise patterns between 2016 and 2021 Census at the provincial level.

```{r temporal-comparison, eval=run_cancensus_examples}
# Get provincial data for 2016 and 2021
# Note: Variable IDs differ between Census years

# 2021 data
prov_2021 <- get_census(
  dataset = "CA21",
  regions = list(C = "01"),
  vectors = c(total_pop = "v_CA21_1085", low_income = "v_CA21_1090"),
  level = "PR",
  geo_format = NA,
  labels = "short",
  quiet = TRUE
)
prov_2021 <- prov_2021[!is.na(prov_2021$total_pop) & prov_2021$total_pop > 0, ]
prov_2021$census_year <- 2021

# 2016 data (different vector IDs for same concept)
prov_2016 <- get_census(
  dataset = "CA16",
  regions = list(C = "01"),
  vectors = c(total_pop = "v_CA16_2540", low_income = "v_CA16_2555"),
  level = "PR",
  geo_format = NA,
  labels = "short",
  quiet = TRUE
)
prov_2016 <- prov_2016[!is.na(prov_2016$total_pop) & prov_2016$total_pop > 0, ]
prov_2016$census_year <- 2016

# Compute surprise for each year
surprise_2021 <- surprise(
  prov_2021,
  observed = low_income,
  expected = total_pop,
  models = c("uniform", "baserate", "funnel")
)
prov_2021$signed_surprise <- get_surprise(surprise_2021, "signed")

surprise_2016 <- surprise(
  prov_2016,
  observed = low_income,
  expected = total_pop,
  models = c("uniform", "baserate", "funnel")
)
prov_2016$signed_surprise <- get_surprise(surprise_2016, "signed")

# Compare changes
comparison <- merge(
  prov_2016[, c("GeoUID", "Region Name", "signed_surprise")],
  prov_2021[, c("GeoUID", "signed_surprise")],
  by = "GeoUID",
  suffixes = c("_2016", "_2021")
)
comparison$change <- comparison$signed_surprise_2021 - comparison$signed_surprise_2016
print(comparison[order(abs(comparison$change), decreasing = TRUE), ])
```

Note: This example is not evaluated because it requires multiple API calls.
Run locally with your CensusMapper API key.

***

# Example 6: Custom Model Space for Census Data

For Census data with known population distributions, create a custom model space.

```{r custom-models, eval=run_cancensus_examples}
# Using the Vancouver housing data
# Create a custom model space

# Create custom model space with uniform and baserate models
# (funnel is better suited for cross-regional comparisons, not local CMA analysis)
housing_models <- model_space(
  bs_model_uniform(),
  bs_model_baserate(housing_data$total_hh)
)

print(housing_models)

# Use surprise() with custom model space
custom_result <- surprise(
  housing_data,
  observed = owner_hh,
  expected = total_hh,
  models = housing_models
)

housing_data$custom_surprise <- get_surprise(custom_result, "signed")

# Summary statistics
cat("Custom model surprise summary:\n")
summary(housing_data$custom_surprise)
```

***

# Interpreting Surprise Values

Understanding what surprise values mean is crucial for proper interpretation:

| Surprise Magnitude | Interpretation | Meaning |
|-------------------|----------------|---------|
| < 0.1 | Trivial | Little update from prior to posterior model probabilities |
| 0.1 - 0.3 | Minor | Small model update |
| 0.3 - 0.5 | Moderate | Noticeable model update |
| 0.5 - 1.0 | Substantial | Strong model update |
| > 1.0 | High | Very strong model update under the chosen model space |

**Key insight**: Surprise measures how much the data discriminates between models, not
just how far a rate is from average. A region with an unusual rate but average population
may show low surprise because all models predict similar things for it. Conversely,
a large region with a rate close to average may show high surprise if it strongly
confirms or refutes the funnel model's predictions about sampling variation.

```{r interpret-surprise, eval=run_cancensus_examples}
# Categorize regions by surprise level
income_data <- income_data %>%
  mutate(
    surprise_category = cut(
      abs(signed_surprise),
      breaks = c(0, 0.1, 0.3, 0.5, 1.0, Inf),
      labels = c("Trivial", "Minor", "Moderate", "Substantial", "High"),
      include.lowest = TRUE
    )
  )

# Summary by category
cat("Census Divisions by surprise level:\n")
print(table(income_data$surprise_category))
```

***

# Session Info

```{r session-info}
sessionInfo()
```
